---
title: 'Working with Data Bases in R: SQL'
author: Nicolas Mollier
date: '2021-01-18'
slug: working-with-data-bases-in-r-sql
categories: []
tags:
  - SQL
  - Query
  - Database
subtitle: ''
summary: ''
authors: []
lastmod: '2021-01-18T10:02:38+01:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
output:
  html:
    toc: true
---

This analysis was part of an assignment in the course Data Science Project Management of the Master Program Data Science in Business and Economics at the University of Tuebingen. It deals with how to establish a connection to a SQL database from within R, writing SQL queries and loading the data into our local memory.

```{r echo = FALSE}
rm(list = ls())
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Root directory and Packages

```{r Set Path to Data}
knitr::opts_knit$set(root.dir = "/home/nicolas/Desktop/Uni/WiSe20_21/DS400_DS_Project_Management/Assignments/Assignment_3")

setwd("/home/nicolas/Desktop/Uni/WiSe20_21/DS400_DS_Project_Management/Assignments/Assignment_3")


```

```{r Packages, message = FALSE, warning = FALSE}

if (!require("microbenchmark")) {
  install.packages("microbenchmark")
}
if (!require("knitr")) {
  install.packages("knitr")
}
if (!require("RSQLite")) {
  install.packages("RSQLite")
}

library(microbenchmark)
library(tidyverse)
library(knitr)
library(DBI)
library(RSQLite)
```

# 2. SQLite database management - soccer data

In a first step I unzip `soccer.zip` in my working directory.

```{r}
unzip("soccer.zip")
```

I create a connection to the database `soccer.sqlite`.

```{r}
con <- dbConnect(RSQLite::SQLite(),
                 dbname = "soccer.sqlite")
```

I load 30 rows of `Match` to get a first impression of the table.

```{r}
match_30 <- dbGetQuery(con, "SELECT * FROM Match LIMIT 30")
glimpse(match_30)
```


## a) Team IDs of FC Bayern Munich and Borussia Dortmund

In a SQL query I select the two columns `id` and `team_long_name` from the `Team` table while restricting the returned cases to those where the variable `team_long_name` takes the values "FC Bayern Munich" or "Borussia Dortmund. 

```{r}
team_id_fcb_bvb <- dbGetQuery(con, "SELECT id, team_long_name
                                    FROM Team 
                                    WHERE team_long_name IN ('FC Bayern Munich', 'Borussia Dortmund')")
team_id_fcb_bvb %>% 
  knitr::kable(align = "l")

```


## b) Match ID of the match in Munich between FC Bayern Munich and Borussia Dortmund in the season of 2015/2016

To get the match id that we are looking for, I have to specify in the WHERE clause the two teams and the season. Since I want to use the results from a), I have to use `paste()` in order to be able to use the team ids that we have already queried above. I also filter for those matches that where in 2015.

```{r}
match_fcb_bvb <- dbGetQuery(con, 
                            paste("SELECT 
                                      Match.id AS match_id, home_team_id, away_team_id, date
                                   FROM Match
                                   WHERE 
                                      home_team_id =", 
                                      # use the id of FCB
                                      filter(team_id_fcb_bvb, team_long_name == "FC Bayern Munich")["id"],
                                      "AND away_team_id =", 
                                      # use the id of BVB
                                      filter(team_id_fcb_bvb, team_long_name == "Borussia Dortmund")["id"],
                                      # filter for those matches that where in 2015
                                      "AND date LIKE '2015%'"
                            )
)

match_id <- match_fcb_bvb %>% 
  select(match_id) %>% 
  as_vector()
```

The match id is `r match_id`.

```{r include = FALSE}

match_fcb_bvb <- dbGetQuery(con, "SELECT Match.id AS match_id, home_team_id, away_team_id, date
                                  FROM Match
                                  WHERE home_team_id = 9823 AND away_team_id = 9789 AND date LIKE '2015%'")
match_id_fcb_bvb <- match_fcb_bvb %>% 
  select(match_id) %>% 
  as_vector()

```


## c) Date of the match between FC Bayern Munich and Borussia Dortmund in the season 2015/2016 

In the query for b) I already selected the column `date`. Therefore, I can use `match_fcb_bvb` here. The `date` column contains the date and the time. Since we are only interested, I remove the time component by using a regular expression.

```{r}
match_fcb_bvb <- match_fcb_bvb %>% 
  mutate(date = str_remove(date, "\\s\\d{2}:\\d{2}:\\d{2}"))

match_fcb_bvb %>% 
  knitr::kable()

# date of the match
match_date <- match_fcb_bvb %>% 
  select(date) %>% 
  as_vector() 
```

The match took place on `r match_date`.

## Alternative to solution a), b), c)

The tasks in a), b), c) could alternatively be solved with the following query that entails two left joins of the tables `Match` and `Team`.

```{r}
solution_abc <- dbGetQuery(con, "SELECT 
                                    match.id, 
                                    home.team_long_name AS home_team_name,
                                    away.team_long_name AS away_team_name, date
                                 FROM Match AS match
                                 LEFT JOIN Team AS home
                                    ON match.home_team_id = home.id
                                 LEFT JOIN Team AS away
                                    ON away_team_id = away.id
                                 WHERE 
                                    home_team_name = 'FC Bayern Munich' AND 
                                    away_team_name = 'Borussia Dortmund' AND
                                    season = '2015/2016'")
solution_abc %>% 
  glimpse()
```


## d) Home team players

Actually, I wanted to use the query that leads to `match_df2`, amend it with UNPIVOT and then join the resulting table with the `Player` table. UNPIVOT was intended to gather the 11 home_player columns into rows while creating a column containing the player ids. This would have enabled the final join with `Player` on id. Unfortunately UNPIVOT did not work. Therefore, I performed the gathering with `dplyr::gather` after having created `match_df2` per SQL query.

```{r}

# the two joins with the Home table give us the team names on which we filter in the WHERE clause
match_df2 <- dbGetQuery(con, 
                        "SELECT 
                            match.*,
                            home.team_long_name AS home_team_name,
                            away.team_long_name AS away_team_name
                         FROM Match AS match
                         LEFT JOIN Team AS home 
                            ON match.home_team_id = home.id
                        LEFT JOIN Team AS away
                            ON away_team_id = away.id
                        WHERE
                           home_team_name = 'FC Bayern Munich' AND
                           away_team_name = 'Borussia Dortmund' AND
                           season = '2015/2016'")

match_df2 %>% 
  glimpse()

# transform match_df2 from wide to long 
# gather the 11 home_player columns into a column called "home_player" and a
# column containing the ids called "home_player_id"
match_df2_gathered <- match_df2 %>%
               gather(key = "home_player",
                     value = "home_player_id",
                     home_player_1 : home_player_11) %>% 
               # remove the players from BVB, date, league_id and country_id etc.
               # since they are of no concern here
               select(-(away_player_1 : away_player_11), 
                      -date,
                      -league_id,
                      -country_id,
                      -home_team_goal,
                      -away_team_goal)
match_df2_gathered %>% 
  knitr::kable()

```

Next, I load the `Player` table in R as `players` and perform the semi join with `match_df2_gathered` in order to get the player names who played in the match for FC Bayern Munich. Those players are saved in a data frame called `match_players`.

```{r}
players <- dbGetQuery(con, paste("SELECT * 
                                  FROM Player"))

# only keep those players that played in that game
match_players <- players %>%
  semi_join(match_df2_gathered,
             by = c("id" = "home_player_id"))

match_players %>% 
  knitr::kable()
```

## e) Average height and weight of the players

For this task we can use `match_players`. 

```{r}
avg_height_weight <- match_players %>%
  select(height, weight) %>% 
  colMeans()
```

Taking the mean for the columns `height` and `weight` results in a mean weight of `r round(avg_height_weight["weight"], 2)` and a mean height of `r round(avg_height_weight["height"],2)`.

## f) Different leagues represented in the table Match

Using the SELECT DISTINCT command in combination with a left join on league id, gives us the 10 distinct leagues from the table `League`.

```{r}

# get all leagues in the table Match
# left join on League gives us the league names
leagues <- dbGetQuery(con, "SELECT DISTINCT league_id, League.* 
                            FROM Match
                            LEFT JOIN League 
                            ON Match.league_id = League.id") 
leagues %>% 
  knitr::kable()
```

